% OMOP analytics efficiency 
The OMOP model is very powerful because it allows a broad spectrum of analysis 
from specialized local models to evidence-based statistical analysis in an 
easy-to-learn and accessible format.

% best of two worlds: features & derived values
As we have seen, the effectiveness of the OMOP model has some weaknesses because 
it seems to focus on consistency rather than performance. However, we have shown 
that it is easy to overcome problems and improve OMOP with a set of design or 
technology optimization and a dedicated structure that ultimately remains a 
standard and shareable because it derives from the original model.

% OMOP transformation
To have such analysis power has a cost of transformation and prior maintenance. 
The transformation of MIMIC into OMOP has required efforts that remain reasonable. 
It is and always will be a work in progress because standard concept mapping is 
an almost infinite process with constant improvements. Fortunately, the published 
version is search-ready and already offers the same scope of data as the original 
MIMIC version and even more with the derived data.


% OMOP contributions
Compared to the original MIMIC data model, working on OMOP offers the ability to 
write standard code and analyses that could benefit other users internationally. 
The MIMIC-OMOP database is available online on physionet as well as the original 
MIMIC database. All existing works are publicly available on github 
\cite{mimic-omop-github} and have been designed to be easily revised, copied or 
enriched according to the OMOP or MIMIC philosophy by any end user who knows SQL.



% Further work
Future work on the evaluation of existing concept mapping through practical 
research studies on local and standard coding will be carried out. In addition, 
we plan to enhance the USAGI OHDSI concept mapping tool to allow the international 
concept mapping suggestion to transform other foreign ICU databases. Finally, 
research on how to articulate FHIR and OMOP to get the best of both worlds 
(information at the patient level versus information at the multi-center level) 
and improve near- bedside care will be done.
