%
% ALTERNATE STORY
%
% 1. ICU benefits from BIG DATA
% 2. MIMIC is the only open access but while very cool it has limitation 
% 3. CDM overcomes MIMIC limitations
% 4. I2B2, FHIR and OMOP CDM's
% 5. OMOP looks better
% 6. Objective threefold are: 1. transform, 2. contribute 3. analysis

% ICU: a paradox 
Intensive Care Units (ICUs) are particularly sensitive units where demand of
care is rising\cite{angus2000} and mortality is up to 30\% which is a major
health care problem \cite{icu-mortality}. Studies have shown that intensivists
use a limited number of clinical information concepts to guide the
decision\cite{icu-evidence} and that the medical practices are sparse and
variable. Knowing that ICU patients' health record are highly detailed
including connected devices this is a paradox.
% Medical database hopes 
The increasing adoption of Electronic Health Records (EHR) systems worldwide
makes it possible to capture large amounts of clinical data
\cite{bigdata-promise} and big data mining has the potential to play an
important role in clinical medicine \cite{bigdata-mining}. Indeed on the basis
of broad patient medical informations expectations are to improve clinical
outcomes and practices, allow personalized medicine and guide decision thought
early warning systems, and also enrol large and multicentric cohort easily
while minimizing costs.
\\
%% Existing ICU databses
%By now several commercial or noncommercial, opensource or nonopensource ICU
%databases have been developed.
%\emph{CUB-REA} is a semi-automatic collection of 25 years and over 300k ICU
%patients stays from 30 distinct ICUs in Paris region and covers administrative
%data only \cite{cubrea-descr,cubrea-website}. Yet it is subject of 15
%international publication and access is under conditions.
%\emph{OutcomeREA} is a manual collection of 20 years over 20k UCU patients
%stays with medium granularity data from 20 distinct ICUs in France. Yet it is
%subject of 50 international publications and access is under conditions.
%\emph{eICU} \cite{eicu-website} is an automatic collection of 2 years of 1.5M
%ICU patients stays from 500 ICU in the USA and covers high granularity
%data \cite{mimic-i2b2}. It is a commercial database.
%\emph{MIMIC} (Medical Information Mart for Intensive Care) is a
%semi-automatic collection of 10 years and over 60k ICU patients stays with very
%high granularity (including EKG) from 1 ICU in Boston. Yet it is subject of 300
%international publication. It is a freely-available database.
%All those databases have their own dedicated model. Their structural model
%are all based on relational database but all do have tables and columns with
%different meaning and granularity. As an example MIMIC do have two inputevents
%tables reflecting its source center changed its EHR. Also their conceptual
%model are mostly different. For example MIMIC do have ICD9 for condition
%terminology, while CUB-REA do have both CIM9 and CIM10. 

% MIMIC database
% - local terminologies
% - one center
% - two successive critical care information systems (CCI)
% - a model that reflect the EHR
\emph{MIMIC} (Medical Information Mart for Intensive Care) is a semi-automatic
collection of 10 years and over 60k ICU patients stays with very high
granularity (including EKG) from two successive critical care information
systems (CCI) of the Beth Israel Deaconess Medical Center in Boston. It is the
first freely-available ICU database and yet it is subject of 300 international
publication. However its monocentric nature makes difficult to generalize the
conclusion on other ICUs. The MIMIC relational data model reflects the originals
CCI as evidenced by the two distinct inputs tables \textit{inputevent\_mv} and
\textit{ouputevent\_cv} \cite{mimic-nature} or the two distinct terminologies
for the physiological data. This leads analysts (data-scientitsts,
statisticians\ldots) to reconcile this heterogeneity during the preprocessing
for each studies.
\\

% Why a Common Data Models
Some studies have shown that using a common data model (CDM) by generalizing
the database on structural (data model) and conceptual (terminology model)
design enables multicentric research, mining rare disease and catalyses
research by allowing sharing practices, source code and
tools \cite{cdm-review,data-enclave}. 
% Criticism for structural model (IMI, FHIR, loose of information)
% Criticism for conceptual model (OMOP standard, FHIR)
However some studies have shown that results are not fully replicable from one
to the other CDM \cite{cdm-comparison} or from one to another center
\cite{omop-replicability}. Lighter approaches argue that keeping the local
conceptual model \cite{fhir-deep} or both original conceptual and structural
\cite{imi-protect} of the database for research leads to better outcomes. On
the first hand keeping MIMIC on its specific form won't solve the limitation of
the previous paragraph and on the other hand a fully standardized form would
introduce other drawbacks. The ideal solution is likely in-between to allow
both local or standardized analysis depending on the research question.
\\

% Existing CDM
% I2B2
Several studies have previously applied CDM to MIMIC. Chronaki et al.
\cite{mimic-i2b2} have evaluated i2b2 to federate MIMIC and eICU. i2b2 is a
medical cohort discovery tool used in more than 200 hospitals over the world.
% i2b2 structural model
The i2b2' highly flexible structural model is able to ingest various domains
and data types in a unique star-schema composed of 5 tables. While i2b2 is
highly efficient for cohort discovery, it's model wasn't designed for ad-hoc
analysis.
% i2b2 conceptual model
The i2b2' conceptual model is a path based table allowing to store any simple
or hierarchical local terminology (graph are not supported). The SHRINE project
goes a step further by proposing a set of standard terminology to federate
multiple i2b2 instances. Each instance needs to map all its local concepts to
the standard's one and vice-versa \cite{shrine-design} (bi-directional
mapping).
% FHIR

More recently Rajkomar et al. have been using \emph{FHIR} (Fast Healthcare
Interoperability Resources) as a CDM to apply deep learning on medical records.
FHIR is a medical data exchange API specification. FHIR provides a structural
CDM that can be materialized as JSON, XML or RDF format. FHIR is flexible and
does not specify a standard conceptual model so that each hospital can add
extension to implement specific data or share within it's local terminology
making each FHIR implementation sensibly divergent. While some research show it
as a promising CDM for ad-hoc analysis \cite{fhir-google} or cohort discovery
% ADD FHIR weakness: loss of data minus raw information
\cite{fhir-paris}, its graph nature adds a layer of transformation making usage
complicated for analysts. 
% fhir structural model
% fhir conceptual model

\textit{OMOP} (Observational Medical Outcomes Partnership Common Data
Model) is a CDM originally designed for multicentric Drug adverse Event and now
extends to medical, clinical and also genomic use cases. OMOP provides both
structural and conceptual model such SNOMED for diagnoses, RxNORM for drug
ingredients and LOINC for laboratory results. While OMOP has proven its
fiability \cite{omop-eval} the fact that concept mapping process is known to
have impact on results \cite{omop-concept-impact} and that applying the same
protocol on different data sources leads to different results
\cite{omop-replicability} reveals the importance of keeping the local codes to
allow local analysis. Several example of transforming databases into OMOP have
been published \cite{omop-german,omop-nashville} and yet OMOP stores 682 milion
patients records from all over the world\cite{omop-bigboy}. OMOP had 5
versions, and prones its strong backward compatibility.
% OMOP
% omop structural model
OMOP structural model consists of several groups of thematic tables (clinical,
costs, derived\ldots). Each clinical domains are stored in various dedicated
tables.
% omop conceptual model
OMOP conceptual model is based on a closure table pattern \cite{closure-table}
able to ingest any simple, hierarchical and also graph terminologies such
SNOMED-CT used locally. In addition to the local terminologies OMOP specifies
and maintains a set of standard terminologies to be mapped uni-directionally
(local to standard) by implementers.
Previous preliminary work have been made on translating MIMIC into OMOP
\cite{mimic-omop-previous} and still work remains to be affined and upgraded to
be furthermore evaluated.
\\

%A dozen of medical CDM exist and a recent study compared their quality
%\cite{} through various characteristics presented in table x.
%From all the candidates, OMOP fits better the criteria. 
%Among the other CDM, OMOP looks like the best fit as it allows both
%multicentric standardised analysis as well as monocentric specific modeling and
%analysis. 

% CDM Comparison
In a recent CDM comparison study \cite{cdm-review,omop-vs-pcornet} OMOP
performs better in the evaluation database criteria compared with the other
models (and PCORnet in particularly) : completeness, integrity, flexibility,
simplicity of integration, and implementability, accommodate the broadest
coverage of standard terminologies, provides more systematic analysis with
analytic library and visualizing tools, provides easier SQL models.
%compared to i2b2
Compared to i2b2/SHRINE, the unilateral mapping methodology of OMOP is more
effective than the bidirectional \cite{shrine-design} SHRINE mapping. Hence
OMOP proposes a broader set of standardised concepts. In terms of structural
CDM OMOP is highly rigourous in how data shall be loaded in a particular table
when i2b2 is highly flexible with a one general table that solution every data
domains. This rigourous approach is necessary for standardisation. Previous
work have loaded i2b2 with MIMICiii \cite{mimic-i2b2} - however, the concept
mapping step have limited the results since i2b2 design do not store the local
ontology or information where OMOP design allows to keep the concept mapping
unfinished. OMOP has this advantage to keep the terminology mapping step not
mandatory by keeping the local codes in a usable format.
%compared to FHIR
As compared to FHIR, OMOP performs better as a conceptual CDM as FHIR does not
specify which terminology to implement. In terms of structural CDM OMOP
relational model can be materialized into csv format and stored in any
relational database when FHIR materialised as json needs some processing and
more skills to be exploited. OMOP shares the advantages of all above models. 
\\

% CHOICE (OMOP & MIMIC)
% choice of mimic
%To demonstrate the feasibility and the opportunity to use a CDM for ICU
%databases, we choose MIMIC since it is freely accessible and then the result of
%this work will be accessible for subsequent improvement, analysis, and
%demonstration. Secondly, MIMIC does have the most broad and high granularity
%database leading to evaluate the ability of the CDM to ingest such complex
%dataset. More importantly, MIMIC stated they would think of a
%CDM\cite{mimic-nature}, that would solve some structure and conceptual
%weaknesses of MIMIC mentioned above and this work should help MIMIC in it's
%goal to build a international ICU database \cite{mimic-i2b2}.
% choice of omop

% Evaluation of MIMIC-OMOP
In order to evaluate the MIMIC to OMOP transformation we propose to answer to
remaining question such the difficulty of transforming/maintaining an OMOP
dataset from an existing one, how well the initial dataset is integrated and
how much data is lost in the process, how clear and simple the model is to be
queried simply and efficiently by analysts, how well design it is to be
enriched by collaborative work, and finally in what extend OMOP can integrate
and makes feedbacks to intensivists in a realtime context.  This work is then
evaluated through 3 axes: Transformations, Contributions and Analytics.
% contribution
The \emph{first} major contribution of this study is to evaluate OMOP in a real
life and well known freely accessible database. The \emph{second} major
contribution is to provide a freely accessible dataset in the OMOP format that
might be useful for researchers. The \emph{third} major contribution is to
provide the OMOP community some useful transformations dedicated to ICU and
that can be reused on any OMOP dataset.
